---
title: "Data analysis exercises for the Learning and Instruction Lab retreat"
format: html
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(lme4)
library(broom.mixed)
library(here)
```

## Before you start

### How these exercises work

Work in pairs.
One person searches for help — using cheat sheets, documentation, or a generative AI tool.
The other person types the code.
Discuss together, but do not copy and paste.
Writing code out by hand, even when you know the answer, is how it sticks.

### Keyboard shortcuts

| Action | Windows | Mac |
|---|---|---|
| Insert code chunk | Ctrl + Alt + I | Cmd + Option + I |
| Insert pipe `|>` | Ctrl + Shift + M | Cmd + Shift + M |
| Run current line or selection | Ctrl + Enter | Cmd + Enter |
| Run current chunk | Ctrl + Shift + Enter | Cmd + Shift + Enter |

### Key tidyverse verbs

These are the main functions you will need today.
If you get stuck, the [tidyverse cheat sheets](https://posit.co/resources/cheatsheets/) are a good first stop.

| Function | Package | What it does |
|---|---|---|
| `pivot_longer()` | tidyr | Reshapes data from wide to long format |
| `mutate()` | dplyr | Creates or modifies columns |
| `filter()` | dplyr | Subsets rows by a condition |
| `select()` | dplyr | Subsets columns |
| `summarise()` | dplyr | Aggregates data, e.g. computing means |

Note: you may see `group_by()` paired with `summarise()` in older code and cheat sheets.
The modern approach uses the `.by` argument inside `summarise()` directly, which is what we use here.

### Pipes and layers: the same idea twice

Two bits of syntax will do a lot of work today.

The **pipe** `|>` chains data through a sequence of steps.
Each step takes the output of the previous one and passes it to the next function.
Think of it as "and then":

```r
data |>
  filter(group == "embodied") |>
  summarise(mean_correct = mean(correct), .by = pid)
```

The **plus sign** `+` in ggplot2 works the same way, but for building plots.
Each `+` adds a new layer on top of the previous one:

```r
ggplot(data, aes(x = group, y = correct)) +
  geom_violin() +
  geom_jitter() +
  labs(x = "Group", y = "Proportion correct")
```

Same mental model, different context: one builds a data pipeline, one builds a plot.

---

## Simple data

Load the simple dataset and create factors.
This code is provided — run it and inspect the result before moving on.

```{r}
#| label: load-simple

dat_simple_wide <- read_csv(
  here("data", "sim_data_simple_wide.csv"),
  show_col_types = FALSE
) |>
  mutate(
    pid = factor(pid),
    group = factor(group, levels = c("control", "embodied"))
  )
```

Take a look at the data.

```{r}
#| label: glimpse-simple

glimpse(dat_simple_wide)
head(dat_simple_wide)
```

---

## Exercise 1: Wide to long

The item columns (`item_1`, `item_17`, etc.) each represent one observation of the same variable: whether a participant got an item correct or not.
In tidy data, each variable has its own column.
So these 8 columns should become two: one for item identity, one for the response.

Your job is to reshape the data from wide to long format using `pivot_longer()`.
The result should have one row per participant per item, with columns `pid`, `group`, `pre_score_z`, `item`, and `correct`.

```{r}
#| label: pivot-simple

# pivot the item columns to long format
# hint: use starts_with() to select the item columns
# hint: names_to is the new column for item names, values_to is the new column for responses
dat_simple_long <- dat_simple_wide |>
  pivot_longer(
    cols = ,
    names_to = ,
    values_to = 
  )
```

Check the result looks right.

```{r}
#| label: check-pivot-simple

# how many rows do you expect? 108 participants x 8 items = 864 rows
# check with nrow() or glimpse() or head()

```

> **Reflect:** why is the long format better suited for plotting and modelling than the wide format?

---

## Exercise 2: Plot the data

Plot the distribution of proportion correct by group.
First summarise to one proportion correct value per participant, then plot.

```{r}
#| label: plot-simple

# hint: summarise prop correct by c(pid, group) first, then pipe to ggplot
# hint: use geom_violin() for the distribution and geom_jitter() for individual points

dat_simple_long |>
  summarise(prop_correct = , .by = c()) |>
  ggplot(aes(x = , y = , fill = , colour = )) +
  geom_violin() +
  geom_jitter() +
  labs(
    x = ,
    y = ,
    title = 
  )
```

> **Reflect:** does the plot suggest a group difference?
> What are the limitations of this summary view?

---

## Exercise 3: Single-level model

Fit a single-level logistic regression predicting `correct` from `group` and `pre_score_z`.
This model ignores the fact that observations are nested within participants.

Use `glm()` with `family = binomial`.
Use `tidy()` from `broom.mixed` to view the output, with `conf.int = TRUE` to get confidence intervals.

```{r}
#| label: glm-simple

# fit a single-level logistic regression
# what goes on the left of ~?
# what predictors go on the right?
m_single <- glm(
  ,
  data = dat_simple_long,
  family = binomial
)

tidy(m_single, conf.int = TRUE)
```

> **Reflect:** what does the coefficient for `group` tell you?
> What is the model assuming about participants that might not be realistic?

---

## Exercise 4: Multilevel model

Now fit a multilevel logistic regression using `glmer()` from the `lme4` package.
Add a varying intercept for each participant.
This allows each person to have their own baseline response probability, over and above the group average.

A note on convergence: `lme4` fits models using maximum likelihood estimation, and multilevel models — especially with binary outcomes — frequently fail to converge.
This is not a bug; it reflects a genuine limitation of the algorithm.
Bayesian software such as `brms` fits the same models via a different algorithm that is more robust and does not have this problem, which is why many researchers prefer it for multilevel modelling.
For the models we are fitting today, `lme4` should be fine.

```{r}
#| label: glmer-simple

# fit a multilevel logistic regression with varying intercepts by pid
# the fixed effects are the same as Exercise 3
# hint: random effects are specified as (1 | grouping_variable)
m_multi <- glmer(
  ,
  data = dat_simple_long,
  family = binomial
)

tidy(m_multi, conf.int = TRUE)
```

> **Reflect:** how does the estimate for `group` change compared to the single-level model?
> What is the varying intercept capturing?

---

## Full data (extension)

If you have time, apply the same workflow to the more complex dataset.
This data includes two test timepoints — post-test and a one-week delay — which motivates a richer model with varying slopes as well as varying intercepts.

Load the full dataset and create factors.
This code is provided — run it and inspect the result before moving on.

```{r}
#| label: load-full

dat_full_wide <- read_csv(
  here("data", "sim_data_full_wide.csv"),
  show_col_types = FALSE
) |>
  mutate(
    pid = factor(pid),
    group = factor(group, levels = c("control", "embodied")),
    test_time = factor(test_time, levels = c("post", "delay"))
  )
```

Take a look at the data.

```{r}
#| label: glimpse-full

glimpse(dat_full_wide)
head(dat_full_wide)
```

---

## Exercise 5: Wide to long (full data)

The structure is the same as before, but now there is an additional column: `test_time`.
Pivot the item columns to long format as you did in Exercise 1.

```{r}
#| label: pivot-full

dat_full_long <- dat_full_wide |>
  pivot_longer(
    cols = ,
    names_to = ,
    values_to = 
  )
```

---

## Exercise 6: Plot the data (full data)

Plot the distribution of proportion correct by group and test time.
Summarise to one proportion correct value per participant per test time first.
Use `geom_violin()` and `geom_jitter()` as before.
Use `facet_wrap()` to show the two test times in separate panels.

```{r}
#| label: plot-full

# hint: summarise by c(pid, group, test_time)
# hint: facet_wrap(~ variable) splits the plot into panels by a variable

dat_full_long |>
  summarise(prop_correct = , .by = c()) |>
  ggplot(aes(x = , y = , fill = , colour = )) +
  geom_violin() +
  geom_jitter() +
  facet_wrap(~ ) +
  labs(
    x = ,
    y = ,
    title = 
  )
```

---

## Exercise 7: Single-level model (full data)

Fit a single-level model predicting `correct` from `group`, `test_time`, their interaction, and `pre_score_z`.
Use `*` to specify an interaction between two predictors.

```{r}
#| label: glm-full

m_single_full <- glm(
  ,
  data = dat_full_long,
  family = binomial
)

tidy(m_single_full, conf.int = TRUE)
```

---

## Exercise 8: Multilevel model (full data)

Fit a multilevel model with varying intercepts and varying slopes for `test_time` by participant.
This allows each person to have their own trajectory across test times, not just their own baseline.

```{r}
#| label: glmer-full

# hint: varying slopes are specified as (1 + slope_variable | grouping_variable)
m_multi_full <- glmer(
  ,
  data = dat_full_long,
  family = binomial
)

tidy(m_multi_full, conf.int = TRUE)
```

> **Reflect:** what does a varying slope for `test_time` capture that a varying intercept alone does not?
> Look at the random effects output — is there meaningful variability in slopes across participants?
